### 2020-05-31 writeup/intro.tex
sec 2

Our second contribution is to use our newly developed technical tool to rigorously explain negative transfer in multi-task learning.
In Section \ref{sec_insight}, we identify three factors including task similarity, data size and covariate shift.
\squishlist
		\item First, we define task dissimilarity as the distance between $\beta_1$ and $\beta_2$ using a simplified isotropic setting of two tasks.
		Using our technical tool, we show a sharp transition from positive to negative transfer as task dissimilarity increases.
		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
		\item Second, we consider the effect of uneven data sizes.
	We begin by observing that adding more labeled data from the source task does not always improve performance.
	Then, we theoretically analyze the benefit of multit-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning, which is a key empirical finding of Taskonomy \cite{ZSSGM18}.
%	We define the \textit{data efficiency ratio} as the smallest $x$ such that if we only use an $$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	%In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.
	%In Section \ref{sec_validate}, we validate that performing multi-task learning can reduce the need for labeled data on 6 sentiment analysis tasks. %, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, we show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, the source task should have the same covariance matrix as the target task to ensure the best performance.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.
\squishend

>>>>

Our second contribution is to use our newly developed tool to rigorously explain negative transfer in multi-task learning.
In Section \ref{sec_insight}, we consider three factors including task model, data size and covariate shift.
\squishlist
		\item First, we define task dissimilarity as the distance between $\beta_1$ and $\beta_2$ using a simplified isotropic setting of two tasks.
		Using our technical tool, we show a sharp transition from positive to negative transfer as task dissimilarity increases.
		Furthermore, we show that negative transfer is more likely to occur when the source task labels are particularly noisy.
%		In Section \ref{sec_validate}, we validate the observation on text and image classification tasks.
%	In , we provide the trade-off between $\norm{\beta_1 - \beta_2}^2$ and a certain function $\Phi(\rho_1, \rho_2)$ to determine the type of transfer.
		\item Second, we consider data size.
	We begin by observing that adding more labeled data from the source task does not always improve performance.
	Then, we analyze the benefit of multi-task learning for reducing the amount of labeled data needed to achieve comparable performance to single-task learning, which is a key empirical finding of Taskonomy \cite{ZSSGM18}.
%	We define the \textit{data efficiency ratio} as the smallest $x$ such that if we only use an $$ fraction of labeled data, then the test error of $\hat{\beta}_t^{\MTL}$ matches that of the $\hat{\beta}_t^{\STL}$ on the entire set.
	%In Proposition \ref{prop_data_efficiency}, we show that the data efficiency ratio of an illustrative example is at most $\alert{\frac {1}{2\rho_t}}$.
	%In Section \ref{sec_validate}, we validate that performing multi-task learning can reduce the need for labeled data on 6 sentiment analysis tasks. %, we observe that just by using \alert{40\%} of the labeled data, the overall accuracy of multi-task learning matches that of learning every task in isolation.
		\item Finally, we show that covariate shift, measured by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$, is another cause for suboptimal performance for $\hat{\beta}_t^{\MTL}$.
		We show that as $n_1 / n_2$ becomes large, the source task should have the same covariance matrix as the target task to ensure the best performance.
%		On the other hand, when $n_1 / n_2$ is small, there are counter examples where having the same covariance matrix is not necessarily the optimal choice.
\squishend

### 2020-05-31 writeup/setup.tex
sec 2

Recall that we have $t$ labeled training datasets, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$, where $X_i\in\real^{n_i\times p}$ and $Y_i\in\real^{n_i}$ for $1\le i\le t$.
%Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Without loss of generality, let the $t$-th task denote the target task.
We consider the following linear multi-task learning architecture.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2,
\end{align}
where $B\in\real^{p\times r}$ and $W_i\in\real^r$.
Here $B$ provides a shared subspace for all tasks and $W_i$ fits $B$ suitably to each task.
Following \cite{WZR20}, we assume that $r < t$, because otherwise minimizing $f(\cdot)$ could result in $BW_i$ to be the single-task optimum.
Hence there is no positive or negative transfer.
We define $\hat{\beta}_t^{\MTL}$ by two steps:
(i) minimizing $f(\cdot)$ over $B$;
(ii) minimize $\set{W_i}_{i=1}^k$ over an independent sample of the training set, e.g. $O(p^{0.99})$ suffices.
For more details, we refer the reader to Appendix \ref{app_proof_sec3}.
For an estimator $\hat{\beta}\in\real^p$, we define the out-of-sample (prediction) loss as
	\begin{align*}
		\te_t(\hat{\beta}) \define \exarg{x}{({x}^{\top}\hat{\beta} - {x}^{\top}\beta_t)^2}
		= (\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t).
	\end{align*}
The single-task estimator $\hat{\beta}_t^{\STL}$ is given by $(X_t^{\top}X_t)^{-1}X_t^{\top}Y_t$.
The bias-variance tradeoff \cite{HTF09} says
	\[ \te_t(\hat{\beta}) =
		\bignorm{\exarg{\varepsilon_t}{\hat{\beta}} - \beta_t}^2 + \exarg{\varepsilon_t}{\bignorm{\hat{\beta} - \exarg{\varepsilon_t}{\hat{\beta}}}^2}. \]
In order to relate the tradeoff to properties of the data, we need tight concentration bounds for the bias and variance.
For this purpose, we consider the high-dimensional regime where $n_i$ is a fixed constant $\rho_i > 1$ times $p$ for every $1\le i\le t$, and $p$ is large.
We focus on a setting where $\rho_t$ is small compared to $\set{\rho_i}_{i=1}^{t-1}$.
This setting captures the need for adding more labeled data to reduce the prediction loss of the target task.
A well-known result for this setting states that $\te_t(\hat{\beta}_t^{\STL}) = \sigma^2 \cdot \tr[(X_t^{\top}X_t)^{-1}\Sigma_t]$ is concentrated around $\frac {\sigma^2} {\rho_t - 1}$ (e.g. Chapter 6 of \cite{S07}), which scales with the data size and noise level of the target task.
However, this result only applies to a single task.
Therefore, our goal is to extend this result to multiple tasks.

@@

As a remark, when there is only the target task, $\rho_1$ equals zero.
Hence $a_2 = 1 - 1/ \rho_2, a_1 = 0$ and Lemma \ref{lem_cov_shift_informal} states that $\tr[(X_2^{\top}X_2)^{-1}\Sigma_2] = \sigma^2 / (\rho_2 - 1)$, which is a well-known result in random matrix theory \cite{S07}.
Hence our result provides a necessary tool to analyze MTL in high dimensions.

>>>>

Recall that we have $t$ labeled training datasets, denoted by $(X_1, Y_1), (X_2, Y_2), \dots, (X_t, Y_t)$, where $X_i\in\real^{n_i\times p}$ and $Y_i\in\real^{n_i}$ for $1\le i\le t$.
%Following \cite{HMRT19,BLLT20}, we assume that for each task $i = 1,2,\dots,t$,  every feature vector is generated as $x = \Sigma_i^{1/2} z$, where $z\in\real^p$ is a random vector with i.i.d. entries of mean zero and unit variance and $\Sigma_i\in\real^{p\times p}$ is a positive semidefinite matrix.
Without loss of generality, let the $t$-th task denote the target task.
We consider the following linear multi-task learning architecture.
\begin{align}
	\label{eq_mtl}
	f(B; W_1, \dots, W_t) = \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2,
\end{align}
where $B\in\real^{p\times r}$ and $W_i\in\real^r$.
Here $B$ provides a shared subspace for all tasks and $W_i$ fits $B$ suitably to each task.
Following \cite{WZR20}, we assume that $r < t$, because otherwise minimizing $f(\cdot)$ could result in $BW_i$ to be the single-task optimum.
Hence there is no positive or negative transfer.
We define $\hat{\beta}_t^{\MTL}$ by two steps:
(i) minimizing $f(\cdot)$ over $B$;
(ii) minimize $\set{W_i}_{i=1}^k$ over an independent sample of the training set, e.g. $O(p^{0.99})$ suffices.
For more details, we refer the reader to Appendix \ref{app_proof_sec3}.
For an estimator $\hat{\beta}\in\real^p$, we define the out-of-sample prediction loss as
	\begin{align*}
		\te_t(\hat{\beta}) = \exarg{x}{({x}^{\top}\hat{\beta} - {x}^{\top}\beta_t)^2}
		= \bignorm{\ex{\hat{\beta}} - \beta_t}^2 + \ex{\bignorm{\hat{\beta} - \ex{\hat{\beta}}}^2},
	\end{align*}
%= (\hat{\beta} - \beta_t)^{\top}\Sigma_t(\hat{\beta} - \beta_t) \\
which can be further decomposed as the bias-variance tradeoff.
The single-task estimator $\hat{\beta}_t^{\STL}$ is given by $(X_t^{\top}X_t)^{-1}X_t^{\top}Y_t$.
In order to relate the tradeoff to properties of the data, we need tight concentration bounds for the bias and variance.
For this purpose, we consider the high-dimensional regime where $n_i$ is a fixed constant $\rho_i > 1$ times $p$ for every $1\le i\le t$, and $p$ is large.
We focus on a setting where $\rho_t$ is small compared to $\set{\rho_i}_{i=1}^{t-1}$.
This setting captures the need for adding more labeled data to reduce the prediction loss of the target task.
A well-known result for this setting states that $\te_t(\hat{\beta}_t^{\STL}) = \sigma^2 \cdot \tr[(X_t^{\top}X_t)^{-1}\Sigma_t]$ is concentrated around $\frac {\sigma^2} {\rho_t - 1}$ (e.g. Chapter 6 of \cite{S07}), which scales with the data size and noise level of the target task.
However, this result only applies to a single task.
Therefore, our goal is to extend this result to multiple tasks.

@@

\textbf{Remark.} When there is only the target task, $\rho_1$ equals zero.
Hence $a_2 = 1 - 1/ \rho_2, a_1 = 0$ and Lemma \ref{lem_cov_shift_informal} states that $\tr[(X_2^{\top}X_2)^{-1}\Sigma_2] = \sigma^2 / (\rho_2 - 1)$, which is a well-known result in random matrix theory \cite{S07}.
Hence our result provides a necessary tool to analyze MTL in high dimensions.
